---
title: "Practicum II CS5200 (part-III)"
author: "Modi, Harshkumar"
date: "Fall 2023"
output: 
  pdf_document
---



```{r - Loading libraries}
# Load the required libraries package
library(RMySQL)
library(ggplot2)
library(scales)

```



```{r - Connecting to the database}
# Connect to the MySQL database
mysql_db <- dbConnect(RMySQL::MySQL(), 
                      host='salesdatabase.ccgfbvk5hpfc.us-east-1.rds.amazonaws.com', 
                      port=3306,
                      username='admin', 
                      password='CS5200Fall23!')

dbExecute(mysql_db, "USE salesdb;")
```

## Analytical Query I: Top five sales reps with the most sales broken down by year.

```{r - Storing query-I output in a dataframe}
query1 <- unique(dbGetQuery(mysql_db, "WITH RankedSales AS (
SELECT year, firstName, lastName, repID, SUM(repQuarterYearSold) as TotalSales,
RANK() OVER (PARTITION BY year ORDER BY SUM(repQuarterYearSold) DESC) as SalesRank
FROM reps_facts
GROUP BY year, repID, firstName, lastName)
SELECT year, firstName, lastName, repID, TotalSales
FROM RankedSales
WHERE SalesRank <= 5
ORDER BY year DESC, SalesRank;
"))
```

```{r - Plotting query-I results}
plot <- ggplot(query1, aes(x=reorder(paste(firstName, lastName, sep=' '), TotalSales), y=TotalSales, fill=factor(year))) +
  geom_bar(stat='identity', position='dodge') +
  theme_minimal() +
  scale_fill_brewer(palette="Pastel1", name="Year") +
  labs(title='Top Five Sales Reps with Most Sales by Year', x='Sales Rep Name', y='Total Sales') +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

print(plot)
```
### The plot above shows the top 5 sales representatives and their sales from each year. The trend we see here is an upward trend in terms of total sales for representatives







## Analytical Query II: Total sold per year per region.
```{r - Storing query-II output in a dataframe}
query2 <- unique(dbGetQuery(mysql_db, "SELECT year, region, SUM(regionTotal) AS Total
                            FROM sales_facts
                            GROUP BY year, region"))
```

```{r - Plotting query-II results}
plot_rotated_x <- ggplot(query2, aes(x=region, y=Total, fill=region)) +
  geom_bar(stat="identity") +
  facet_wrap(~year) + # Facets for each year
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  labs(title="Region Totals by Year", x="Region", y="Total Sales") +
  scale_fill_brewer(palette="Dark2") 

print(plot_rotated_x)

```
### The plot above shows the total sales region wise in each year. 2021 had the most number of sales amongst all of the data




## Analytical Query III: Total sold per quarter per year. 
```{r - Storing query-III output in a dataframe}

query3 <- unique(dbGetQuery(mysql_db, "SELECT quarter, year, SUM(quarterYearTotal) AS total
                            FROM sales_facts
                            GROUP BY quarter, year"))
```

```{r - Plotting query-III results}

query3$year <- factor(query3$year)

ggplot(query3, aes(x=quarter, y=total, group=year, color=year)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(title="Quarterly Sales Total by Year",
       x="Quarter",
       y="Total Sales",
       color="Year") +
  scale_x_continuous(breaks=1:4) + 
  scale_y_continuous(labels=label_number())

```
### The plot above shows the total sales per quarter for all years. The trend of high sales in 2021 is evident once again.


```{r - Disconnecting database}

dbDisconnect(mysql_db)
```


