---
title: "Practicum I CS5200"
author: "Modi, Harshkumar"
date: "Fall 2023"
output: 
  pdf_document
---


```{r Importing required packages for data manipulation and connecting with database server, warning=FALSE}
#install.packages("DBI")
#install.packages("tidyverse")
#install.packages("RMySQL")
#install.packages("dplyr")
library(tidyverse)
library(RMySQL)
library(DBI)
library(dplyr)
```

## Connect to Database
```{r - Creating a connection with the server on AWS}
con <- dbConnect(RMySQL::MySQL(), 
                 host = "myrds.ccgfbvk5hpfc.us-east-1.rds.amazonaws.com", 
                 port = 3306,
                 dbname = "birdStrikesDB",
                 username = "admin", 
                 password = "CS5200Fall23!")
```



## Creating database and tables
```{sql Drops strikes table if exists,connection=con, eval=F}
DROP TABLE IF EXISTS strikes;
```

```{sql Drops conditions table if exists, connection=con, eval=F}
DROP TABLE IF EXISTS conditions;
```

```{sql Drops flights table if exists, connection=con, eval=F}
DROP TABLE IF EXISTS flights ;
```

```{sql Drops airports table if exists, connection=con, eval=F}
DROP TABLE IF EXISTS airports;
```

## Q4 (A) Creating flights table
```{sql Create flights table if exists,connection=con}
CREATE TABLE IF NOT EXISTS flights (
  fid INTEGER NOT NULL AUTO_INCREMENT,
  date DATE NOT NULL,
  origin INTEGER NOT NULL,
  airline TEXT NOT NULL,
  aircraft TEXT NOT NULL,
  altitude INTEGER CHECK (altitude>=0),
  heavy INTEGER NOT NULL,
  PRIMARY KEY (fid)) ;
```

## Q4 (B) Creating airports table
```{sql Create table airports if exists, connection=con}
CREATE TABLE IF NOT EXISTS airports (
  aid INTEGER NOT NULL AUTO_INCREMENT,
  airportName TEXT NOT NULL,
  airportState TEXT NOT NULL,
  airportCode TEXT,
  PRIMARY KEY (aid));
```

## Q4 (C) Adding foreign key origin to reference airports table
```{sql Adds foreign key origin to reference airports table, connection=con}
ALTER TABLE flights ADD FOREIGN KEY (origin) REFERENCES airports(aid);
```

## Q4 (D) Creating table conditions
```{sql Create table conditions if exists, connection=con}
CREATE TABLE IF NOT EXISTS conditions (
  cid INTEGER NOT NULL AUTO_INCREMENT,
  sky_condition TEXT NOT NULL,
  explanation TEXT,
  PRIMARY KEY (cid));
```

## Q4 (E) Creating table strikes
```{sql Create table strikes if exists, connection=con}
CREATE TABLE IF NOT EXISTS strikes (
  sid INTEGER NOT NULL AUTO_INCREMENT,
  fid INTEGER NOT NULL,
  numbirds INTEGER NOT NULL,
  IMPACT TEXT NOT NULL,
  damage INTEGER NOT NULL,
  altitude INTEGER CHECK (altitude>=0),
  conditions INTEGER NOT NULL,
  PRIMARY KEY (sid));
```

## Q4 (F) Adding foreign key fid to reference flights table
```{sql Adds foreign key fid to reference flights table, connection=con}
ALTER TABLE strikes ADD FOREIGN KEY (fid) REFERENCES flights(fid);
```
## Q4 (F) Adding foreign key conditions to reference conditions table
```{sql Adds foreign key conditions to reference conditions table, connection=con}
ALTER TABLE strikes ADD FOREIGN KEY (conditions) REFERENCES conditions(cid);
```

## Q4 (G) Describing each table
```{sql Taking a look at airports table, connection=con, eval=FALSE}
DESCRIBE airports;
```


```{sql Describes the flights table, connection=con, eval=FALSE}
DESCRIBE flights;
```


```{sql Describes the strikes table, connection=con, eval=FALSE}
DESCRIBE strikes;
```


```{sql Describes the conditions lookup table, connection=con, eval=FALSE}
DESCRIBE conditions;
```

## Q5 Importing csv data
```{r Importing csv data}
bds.raw <- read.csv("https://s3.us-east-2.amazonaws.com/artificium.us/datasets/BirdStrikesData-V2.csv", header = TRUE)
#Checking all columns and it's datatype in the csv data
str(bds.raw)
```

## Removing N/A and empty values
```{r - Checking the airports table where origin column is N/A and "" values, eval = T, warning = F}
print(sum(bds.raw$origin == "N/A"))
print(sum(bds.raw$airport == ""))
```

## Filling N/A and empty values with "unknown" as per question
```{r - Fill the airports table with airport "" and origin with N/A to unknown, eval = T, warning = F}
## removing blank and NA values and replacing with "unknown"
bds.raw$airport[bds.raw$airport == ""] <- "unknown"
bds.raw$origin[bds.raw$origin == "N/A"] <- "unknown"
```

## Checking if N/A and empty values still persists
```{r - Checking Origin column with unknown values, eval = T, warning = F}
## checking again to see if blank values are removed
print(sum(bds.raw$origin == "N/A"))
print(sum(bds.raw$airport == ""))
```


## Creating ids for tables according to their schema
```{r Creating ids for tables according to their schema}
bds.raw <- bds.raw %>%
  mutate(
    aid = as.integer(factor(airport, levels = unique(airport))), #aid - creates a unique ID for airports 
    cid = as.integer(factor(sky_conditions, levels = unique(sky_conditions))), #cid - creates a unique ID for conditions ("No Cloud", "Some Cloud", "Overcast")
    fid = seq_along(rid) #fid - creates unique ID for each row
  )
head(bds.raw,5)
```

## Creating dataframe airports
```{r Selecting required rows from the dataframe for the airports table and changes datatype according to our defined sql table, eval = T, warning = F}
#Create the airports table and populate it with unique airport data from the incidents table
airports <- unique(bds.raw[c("aid","airport", "origin")])
# mapping the dataframe airports to the table airports in database
colnames(airports) <- c("aid","airportName", "airportState")
head(airports, 1)
```

## Creating dataframe flights
```{r Selecting required rows from the dataframe for the flights table and changes datatype according to our defined sql table}
flights <- bds.raw %>% select(fid, flight_date, origin = aid, airline, model, altitude_ft, heavy_flag)
colnames(flights) <- c("fid","date", "origin", "airline", "aircraft", "altitude", "heavy")
flights <- flights %>%
  mutate(
    date = as.Date(date, format = "%m/%d/%Y"),  
    altitude = as.integer(gsub(",", "", altitude)),
    heavy = ifelse(heavy == "Yes", 1, 0)         # Converting 'Yes' to 1 and 'No' to 0
  )

#Displays first row of selected rows of dataframe
head(flights,1)
```

## Creating dataframe conditions
```{r Selecting required rows from the dataframe for the conditions lookup table and making changes in datatype according to our defined sql table}
conditions <- bds.raw[c("cid","sky_conditions")]
colnames(conditions) <- c("cid","sky_condition")
conditions <- unique(conditions)
head(conditions,1)
```

## Creating dataframe strikes
```{r Selecting required rows from the dataframe for the strikes table and making changes in datatype according to our defined sql table}
strikes <- bds.raw[c("fid","wildlife_struck", "impact", "damage", "altitude_ft", "cid")]
colnames(strikes) <- c("fid","numbirds", "IMPACT", "damage", "altitude", "conditions")
strikes$damage <- recode(strikes$damage, 
                            `Caused damage` = 1,
                            `No damage` = 0)
strikes <- strikes %>% mutate(
  damage = as.integer(damage),
  conditions = as.integer(conditions),
  altitude = as.integer(gsub(",", "", altitude)))
head(strikes, 1)
```

## Q6 Inserting data (from dataframes) into the sql tables
```{r}
dbWriteTable(con, name = "airports", value = airports, overwrite = FALSE, append = TRUE, row.names = FALSE)
```

```{r Inserting all the rows of the dataframe to the "flights" sql table}
dbWriteTable(con, name = "flights", value = flights, overwrite = FALSE, append = TRUE, row.names = FALSE)
```

```{r Inserting all the rows of the dataframe to the "conditions" sql table}
dbWriteTable(con, name = "conditions", value = conditions, overwrite = FALSE, append = TRUE, row.names = FALSE)
```

```{r Inserting all the rows of the dataframe to the "strikes" sql table}
dbWriteTable(con, name = "strikes", value = strikes, overwrite = FALSE, append = TRUE, row.names = FALSE)
```







## Q7 Showing glimpses of data in each table
```{sql Shows the first 10 rows of aiports table, connection=con}
SELECT * from airports limit 10;
```

```{sql Displaying first 10 rows of the "flights" sql table, connection=con}
SELECT * FROM flights LIMIT 10;
```

```{sql Displaying first 10 rows of the "conditions" sql table, connection=con, eval=FALSE}
SELECT * FROM conditions LIMIT 10;
```

```{sql Displaying first 10 rows of the "strikes" sql table, connection=con, eval=FALSE}
SELECT * FROM strikes LIMIT 10;
```

# Q8 Finding the top 10 states with the greatest number of bird strike incidents
```{sql Finding the top 10 states with the greatest number of bird strike incidents, connection = con}
SELECT 
    a.airportState AS state,
    COUNT(s.sid) AS number_of_incidents
FROM 
    flights f
JOIN 
    airports a ON f.origin = a.aid
JOIN 
    strikes s ON f.fid = s.fid
GROUP BY 
    a.airportState
ORDER BY 
    number_of_incidents DESC
LIMIT 10;
```

# Q9 Finding the airlines that had an above average number bird strike incidents
```{sql Finding the airlines that had an above average number bird strike incidents, connection=con}
WITH AirlineIncidents AS (
    SELECT 
        f.airline,
        COUNT(s.sid) AS incidents
    FROM 
        flights f
    JOIN 
        strikes s ON f.fid = s.fid
    GROUP BY 
        f.airline
),

AverageIncidents AS (
    SELECT 
        AVG(incidents) AS avg_incidents
    FROM 
        AirlineIncidents
)

SELECT 
    ai.airline,
    ai.incidents
FROM 
    AirlineIncidents ai, AverageIncidents av
WHERE 
    ai.incidents > av.avg_incidents
ORDER BY 
    ai.incidents DESC;

```

## Q10 Finding the (total) number of birds that struck aircraft by month
```{r Finding the (total) number of birds that struck aircraft by month}
# Execute the SQL query
query <- "
SELECT 
    EXTRACT(MONTH FROM f.date) AS month, 
    SUM(s.numbirds) AS total_birds_struck
FROM 
    flights AS f
JOIN 
    strikes AS s ON f.fid = s.fid
GROUP BY 
    month
ORDER BY 
    total_birds_struck DESC;
"
#Storing the results of the query in a dataframe
monthlyStrikes <- suppressWarnings(dbGetQuery(con, query))
monthlyStrikes$month <- months(as.Date(paste(monthlyStrikes$month, "1", sep="-"), format="%m-%d"))
monthlyStrikes$month[is.na(monthlyStrikes$month)] <- "Unknown"


#Printing only the first 6 months of data from the dataframe
print(monthlyStrikes[1:6,])
```

## Q11 Building a column chart that plots month along the x-axis versus number of birds on the y-axis
```{r Building a column chart that plots month along the x-axis versus number of birds on the y-axis}
# Assuming df is the dataframe you got from the previous question
monthlyStrikes <- monthlyStrikes[-1,]

# Creating a bar plot without x-axis labels
bp <- barplot(monthlyStrikes$total_birds_struck, 
              names.arg=NULL, # suppress x-axis labels
              main="Bird Strikes by Month",
              xlab="Month",
              ylab="Number of Birds Struck",
              col="skyblue",
              border="black",
              ylim=c(0, max(monthlyStrikes$total_birds_struck) + 1500)) # +10 for a little extra space at the top

# Adding rotated x-axis labels
axis(side=1, at=bp, labels=monthlyStrikes$month, las=2, srt=30, adj=1)

# Adding data labels
if (length(monthlyStrikes$total_birds_struck) > 1) {
  text(x=bp, 
       y=monthlyStrikes$total_birds_struck + 5, # +5 to position the labels just above the bars
       labels=monthlyStrikes$total_birds_struck, 
       cex=0.7, 
       pos=3)
} else {
  text(x=bp[1], 
       y=monthlyStrikes$total_birds_struck + 5, 
       labels=monthlyStrikes$total_birds_struck, 
       cex=0.7, 
       pos=3)
}
```

## Q12 Creating a stored procedure in MySQL that adds a new strike to the database
```{sql Creating a stored procedure in MySQL that adds a new strike to the database, connection=con}
CREATE PROCEDURE IF NOT EXISTS AddBirdStrike(
    -- Parameters for Airports
    IN p_airportName TEXT, 
    IN p_airportState TEXT, 
    IN p_airportCode TEXT,
    
    -- Parameters for Flights
    IN p_date DATE,
    IN p_airline TEXT,
    IN p_aircraft TEXT,
    IN p_altitude INTEGER,
    IN p_heavy BOOLEAN,
    
    -- Parameters for Strikes
    IN p_numbirds INTEGER,
    IN p_impact TEXT,
    IN p_damage BOOLEAN,
    IN p_strike_altitude INTEGER,
    
    -- Parameters for Conditions
    IN p_sky_condition TEXT,
    IN p_explanation TEXT
)
BEGIN
    DECLARE v_origin INTEGER;
    DECLARE v_fid INTEGER;
    DECLARE v_conditions INTEGER;

    -- Check if airport exists, if not insert it
    SELECT aid INTO v_origin FROM airports WHERE airportCode = p_airportCode;
    IF v_origin IS NULL THEN
        INSERT INTO airports(airportName, airportState, airportCode) 
        VALUES (p_airportName, p_airportState, p_airportCode);
        SET v_origin = LAST_INSERT_ID();
    END IF;

    -- Insert flight and get the fid
    INSERT INTO flights(date, origin, airline, aircraft, altitude, heavy) 
    VALUES (p_date, v_origin, p_airline, p_aircraft, p_altitude, p_heavy);
    SET v_fid = LAST_INSERT_ID();

    -- Check if conditions exist, if not insert it
    SELECT cid INTO v_conditions FROM conditions WHERE sky_condition = p_sky_condition AND explanation = p_explanation;
    IF v_conditions IS NULL THEN
        INSERT INTO conditions(sky_condition, explanation)
        VALUES (p_sky_condition, p_explanation);
        SET v_conditions = LAST_INSERT_ID();
    END IF;

    -- Insert strike
    INSERT INTO strikes(fid, numbirds, impact, damage, altitude, conditions)
    VALUES (v_fid, p_numbirds, p_impact, p_damage, p_strike_altitude, v_conditions);
    
END;

```

## Q13 Testing the defined procedure by adding new row of data
```{r Testing the defined procedure by adding new row of data}
sample_airportName <- "Northeastern Airport"
sample_airportState <- "Northeastern Land"
sample_airportCode <- ""
sample_date <- "2023-10-31"
sample_airline <- "NEU"
sample_aircraft <- "Khoury"
sample_altitude <- 10000
sample_heavy <- TRUE
sample_numbirds <- 1
sample_impact <- "minor"
sample_damage <- FALSE
sample_strike_altitude <- 8000
sample_sky_condition <- "Clear"
sample_explanation <- "Normal Condition"

# Call the stored procedure using sprintf
query <- sprintf(
    "CALL AddBirdStrike('%s', '%s', '%s', '%s', '%s', '%s', %d, %d, %d, '%s', %d, '%s', '%s', '%s')",
    sample_airportName, sample_airportState, sample_airportCode, 
    sample_date, sample_airline, sample_aircraft, sample_altitude, as.integer(sample_heavy), 
    sample_numbirds, sample_impact, as.integer(sample_damage), sample_strike_altitude, 
    sample_sky_condition, sample_explanation
)

# Execute the stored procedure
dbSendQuery(con, query)
```

## Verifying the insertion by fetching the latest row in strikes table
```{r Verifying the insertion by fetching the latest row in strikes table}
result <- dbGetQuery(con, "SELECT * FROM strikes ORDER BY sid DESC LIMIT 1")
print(result)
```

## Verifying the insertion by fetching the latest row in flights table
```{r Verifying the insertion by fetching the latest row in flights table}
result <- dbGetQuery(con, "SELECT * FROM flights ORDER BY fid DESC LIMIT 1")
print(result)
```

## Verifying the insertion by fetching the latest row in airports table
```{r Verifying the insertion by fetching the latest row in airports table}
result <- dbGetQuery(con, "SELECT * FROM airports ORDER BY aid DESC LIMIT 1")
print(result)
```

## Disconnecting the database
```{r Disconnect database}
dbDisconnect(con)
```
